The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic

### 2021-05-25 R0-Density-Reanalysis/Cover-Letter-v1.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

Please find attached the paper titled "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic". The paper illustrates the importance of reproducibility in research in the particular context of COVID-19. It takes a recent paper from the literature (Sy et al., 2021, published in PLoS ONE) and demonstrates how openly sharing data and code can accelerate research.

>>>>

Please find attached the paper titled "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic". The paper illustrates the importance of reproducibility in research in the particular context of COVID-19. Research on population density and the spread of COVID-19 has been published in Science of the Total Environment (see Sharifi and Khavarian-Garmsir, 2020; and Sun et al., 2020).

The present submission includes some original research, since in it I reanalyze data from a recent paper (Sy et al., 2021, published in PLoS ONE) and show that after taking into account sample selection the conclusions are drastically different. I think, however, that this paper is better classified as a "Discussion", since the reanalysis is the motivation for a broader discussion about reproducibility in COVID-19 research, and an example of how openly sharing data and code can accelerate research.

### 2021-05-25 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Paper reformatted for submission to Science of the Total Environment after PLoS ONE desk rejection

---
title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    corresponding: paezha@mcmaster.ca
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Given the high stakes of the situation, it is essential that scientific activites, on which good policy depends, are as transparent and reproducible as possible. Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena. In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2. Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequantial changes in the conclusions, as the empirical example in this paper shows.

author_summary: |

bibliography: bibliography.bib
output: rticles::plos_article
csl: plos.csl
---

@@

In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]. The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic. Some important questions hang on the results of these research efforts. For example, are lower density regions safer from the pandemic? Are de-densification policies warranted, even if just in the short term? And in the longer term, will the risks of life in high density regions presage a flight from cities? Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed. Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers. To this end, a recent study by Sy et al. [@Sy2021population] is chosen as an example of reproducible research. The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis. Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors. The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research. This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification. The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].

@@

The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus. It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography @Moore1970urban; @Moore1970some; @Farber2011running]. Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect]. The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].

@@

One salient aspect of the analysis in SWN is that the basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold. As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations. A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated? What are the implications of the spatial sampling framework used in the analysis? Is it possible to spatially interpolate the missing observations? And, was there evidence of spatial autocorrelation in the residuals of the models? These questions are relevant and their implications important. Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis. This means that an independent expert can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.

@@

In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of a spatial modeler are asked, and the data are reanalyzed.

>>>>

---
title: "The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Given the high stakes of the situation, it is essential that scientific activities, on which good policy depends, are as transparent and reproducible as possible. Reproducibility is key for the efficient operation of the self-correction mechanisms of science, which work to weed out errors and refine our understanding of social and physical phenomena. In this paper, the importance of reproducibility is illustrated for the case of the association between population density and the the spread of SARS-CoV-2. Transparency and openness means that the same problem can, with relatively modest efforts, be examined by independent researchers who can verify findings, and bring to bear different perspectives, approaches, and methods—sometimes with consequential changes in the conclusions, as the empirical example in this paper shows.

journal: "Science of the Total Environment"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

\newpage

@@

In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications are the topic of ongoing research [@Sharifi2020covid]. The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, brought a torrent of research into the associations between density and the spread of the pandemic. Some important questions hang on the results of these research efforts. For example, are lower density regions safer from the pandemic? Are de-densification policies warranted, even if just in the short term? And in the longer term, will the risks of life in high density regions presage a flight from cities? Over the past year, numerous papers have sought to throw light into the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed. Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of reproducibility in research, particularly in the context of the flood of COVID-19 papers. To this end, a recent study by Sy et al. [-@Sy2021population] is chosen as an example of reproducible research. The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis. Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors. The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research. This paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification. The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, is available in a public repository^[\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}].

@@

The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus. It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some]. Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect]. The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].

@@

One salient aspect of the analysis in SWN is that the basic reproductive number can only be calculated reliably with a minimum number of cases, and a large number of counties did not meet such threshold. As researchers do, SWN made modelling decisions, in this case basing their analysis only on counties with valid observations. A modeler with expertise in different methods would likely ask some of the following questions on reading SWN's paper: how were missing counties treated? What are the implications of the spatial sampling framework used in the analysis? Is it possible to spatially interpolate the missing observations? These questions are relevant and their implications important. Fortunately, SWN are an example of a reasonably open, reproducible research product: their paper is accompanied by (most of) the data and (most of) the code used in the analysis. This means that an independent expert can, with only a moderate investment of time and effort, reproduce the results in the paper, as well as ask additional questions.

@@

In the following sections, the analysis of RWN is reproduced, some relevant questions from the perspective of an independent researcher are asked, and the data are reanalyzed.
